---
title: "statistics challenge"
author: "Group D. rerio"
date: "2023-09-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, include=F}
## load and install packages

packages <- c("dplyr", "tidyverse")

for (package in packages) {
  if (!require(package, character.only = T, quietly = T)) {
    install.packages(package, repos = "http://lib.stat.cmu.edu/R/CRAN")
  }
  }
for (package in packages) {
  library(package, character.only = T)
}
```

QUESTION: DID MORE COVID-19 RELATED DEATHS OCCUR IN COMMONWEALTH NATIONS THAN
IN THOSE WHO GAINED INDEPENDENCE FROM THE BRITISH EMPIRE ON THE DAY QUEEN ELIZABETH THE SECOND DIED (SEPTEMBER 8TH 2022)?

```{r data}
# load data
data <- read.csv("https://covid.ourworldindata.org/data/owid-covid-data.csv",
                 na.strings = "", header=T)
# Clean data
as_tibble(data)
data_cleaned <- data |>
  select(date,new_deaths,location) |>
  filter(!is.na(new_deaths)) |>
  group_by(location,date) |>
  mutate(death_sum = sum(new_deaths))

# Catagorize commonwealth nations versus nations who gained independence
still_under_their_reign <- c("Australia","United Kingdom")
freed_from_their_majesty <- c("Ghana","Egypt")

# Select all data the day Queen Elizabeth II died
data_from_day_queen_died <- data_cleaned[data_cleaned$date == "2022-09-08",]

# Pull data for each category on subsetted data
independent_deaths_day_queen_died <- data_from_day_queen_died[data_from_day_queen_died$location %in% freed_from_their_majesty,4]

monarchy_deaths_day_queen_died <- data_from_day_queen_died[data_from_day_queen_died$location %in% still_under_their_reign,4]

# Run t test to see if deaths in independent countries are less than deaths in commonwealth countries
t.test(independent_deaths_day_queen_died$death_sum,monarchy_deaths_day_queen_died$death_sum,alternative = "less")
```

p-value = 0.03128
p-value is less than 0.05, so we reject the null hypothesis.
THEREFORE: ON THE DAY THE QUEEN ELIZABETH II DIED, HER COMMONWEALTH CITIZENS SUFFERED MORE COVID-19 RELATED DEATHS THAN THOSE IN NATIONS WHO EARNED THEIR INDEPENDENCE FROM THE BRITISH EMPIRE! We recommend avoiding the passing of heads of state during public health emergencies of international concern.